---
title: "Case study: tracing ADaM back to SDTM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: tracing ADaM back to SDTM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Traceability is a filing requirement, but on most studies it lives as a
spreadsheet someone maintains by hand. `r4subtrace` treats it as data instead:
it takes your ADaM metadata, your SDTM metadata, and a mapping sheet, builds a
graph, and tells you exactly which variables can be traced back to source and
which cannot.

This case study uses the example CDISC pilot metadata that ships with
`r4subdata`.

```{r libs, message = FALSE}
library(r4subtrace)
library(r4subdata)
```

## The inputs

Three tables: what is in ADaM, what is in SDTM, and how the two connect.

```{r inputs}
head(adam_metadata, 3)
head(sdtm_metadata, 3)
head(trace_mapping, 3)
```

The mapping carries a derivation description and a confidence score for each
link, which is what lets the model tell a direct copy apart from a documented
derivation.

## Building the model

`build_trace_model()` turns the three tables into nodes and edges, and flags the
problems it finds along the way.

```{r model}
tm <- build_trace_model(adam_metadata, sdtm_metadata, mapping = trace_mapping)
tm
```

The orphans in that summary are ADaM variables with no mapping row at all. They
are the first thing to chase, because an untraceable variable is exactly what a
reviewer will stop on.

```{r orphans}
tm$diagnostics$orphans
```

## Trace levels

`compute_trace_levels()` assigns each ADaM variable a level from L0 (no mapping,
no derivation) up to L3 (mapped with high confidence or a documented
derivation).

```{r levels}
levels <- compute_trace_levels(tm)
table(levels$trace_level)
```

That distribution is the honest coverage picture. Variables at L0 or L1 are the
gap.

```{r gap}
levels[levels$trace_level < 2, c("adam_dataset", "adam_var", "trace_level")]
```

## Coverage as indicators

`trace_model_to_evidence()` turns the model into R4SUB evidence, and
`trace_indicator_scores()` summarises it into the numbers that feed the Trace
pillar of the Submission Confidence Index.

```{r indicators}
ctx <- r4subcore::r4sub_run_context(study_id = "CDISCPILOT01", environment = "UAT")
ev  <- trace_model_to_evidence(tm, ctx = ctx)
trace_indicator_scores(ev)
```

Coverage, orphan count, ambiguous mappings, and mean trace level: those are the
signals the readiness score reads. Improving them is not a documentation chore
done at the end, it is a number the whole team can watch move as the mapping
gets completed.
