---
title: "Traceability Analysis with r4subtrace"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Traceability Analysis with r4subtrace}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Traceability is a core regulatory requirement: reviewers must be able to
follow every analysis variable from its source SDTM domain through any
derivation to the final ADaM dataset. The `r4subtrace` package builds a
directed trace model and evaluates coverage.

```{r load}
library(r4subtrace)
```

## Default configuration

```{r config}
cfg <- trace_config_default()
str(cfg)
```

## Building a trace model

`build_trace_model()` takes ADaM metadata, SDTM metadata, and an optional
mapping sheet. It returns a `trace_model` object with nodes, edges, and
diagnostic information.

```{r trace-model}
adam_meta <- data.frame(
  dataset  = rep("ADSL", 5),
  variable = c("STUDYID", "USUBJID", "AGE", "SEX", "TRT01P"),
  label    = c("Study ID", "Unique Subject ID", "Age",
               "Sex", "Planned Treatment"),
  stringsAsFactors = FALSE
)

sdtm_meta <- data.frame(
  dataset  = c(rep("DM", 4), "EX"),
  variable = c("STUDYID", "USUBJID", "AGE", "SEX", "EXTRT"),
  label    = c("Study ID", "Unique Subject ID", "Age",
               "Sex", "Treatment Name"),
  stringsAsFactors = FALSE
)

mapping <- data.frame(
  adam_dataset = rep("ADSL", 5),
  adam_var     = c("STUDYID", "USUBJID", "AGE", "SEX", "TRT01P"),
  sdtm_domain  = c("DM", "DM", "DM", "DM", "EX"),
  sdtm_var     = c("STUDYID", "USUBJID", "AGE", "SEX", "EXTRT"),
  stringsAsFactors = FALSE
)

tm <- build_trace_model(adam_meta, sdtm_meta, mapping = mapping)
print(tm)
```

## Inspecting the trace model

The `nodes` tibble lists all assets; `edges` describes the relationships:

```{r nodes-edges}
head(tm$nodes)
head(tm$edges)
```

Diagnostic information flags any orphan variables (unmapped ADaM variables):

```{r diagnostics}
tm$diagnostics$orphans
```

## Computing trace levels

`compute_trace_levels()` summarises coverage per ADaM dataset:

```{r trace-levels}
tl <- compute_trace_levels(tm)
tl
```

## Indicator scores from evidence

If you have an evidence table with `indicator_domain == "trace"`,
`trace_indicator_scores()` computes per-indicator aggregates:

```{r indicator-scores}
ev_trace <- data.frame(
  indicator_id     = c("T-001", "T-001", "T-002"),
  indicator_domain = "trace",
  result           = c("pass", "warn", "fail"),
  metric_value     = c(1.0, 0.8, 0.0),
  metric_unit      = "proportion",
  severity         = c("info", "medium", "high"),
  stringsAsFactors = FALSE
)

trace_indicator_scores(ev_trace)
```

## Partial mapping (orphans)

When a variable has no mapping entry, it appears as an orphan:

```{r orphans}
adam_partial <- data.frame(
  dataset  = rep("ADSL", 3),
  variable = c("USUBJID", "AGE", "DERIVED_VAR"),
  label    = c("Unique Subject ID", "Age", "Derived Variable"),
  stringsAsFactors = FALSE
)
mapping_partial <- data.frame(
  adam_dataset = c("ADSL", "ADSL"),
  adam_var     = c("USUBJID", "AGE"),
  sdtm_domain  = c("DM", "DM"),
  sdtm_var     = c("USUBJID", "AGE"),
  stringsAsFactors = FALSE
)
tm2 <- build_trace_model(adam_partial, sdtm_meta, mapping = mapping_partial)
tm2$diagnostics$orphans
```

Orphaned variables (like `DERIVED_VAR` above) should have derivation text
documented to satisfy U-002 and T-002 requirements.

## Coverage report for the ADRG

Once a trace model is built, `export_traceability_report()` produces the
coverage table an ADRG appendix needs: how much of each ADaM dataset traces
back to SDTM, which variables do not, and an overall summary row.

```{r coverage}
report <- export_traceability_report(tm)
report
```

A variable counts as traced when its trace level reaches `traced_min_level`
(the default of 2 treats an existing SDTM mapping as traced). Raise it to 3 to
require a high-confidence mapping. Pass `format = "csv"` or `format = "xlsx"`
with a `path` to write the table straight into a submission package:

```{r coverage-csv, eval=FALSE}
export_traceability_report(tm, format = "csv", path = "traceability.csv")
```

## Multi-step derivation chains

A direct ADaM-to-SDTM mapping is only the simplest case. Many variables are
derived from other ADaM variables that are themselves derived from SDTM. A
single hop misses those links. `trace_chains()` follows the full path back to
source, using an `adam_derivations` table for the ADaM-to-ADaM steps.

```{r chains}
adam <- data.frame(
  dataset  = c("ADSL", "ADAE", "ADTTE"),
  variable = c("AGE", "AAGE", "TAGE"),
  label    = c("Age", "Analysis Age", "Time-to-event Age")
)
sdtm <- data.frame(dataset = "DM", variable = "AGE", label = "Age")
map  <- data.frame(
  adam_dataset = "ADSL", adam_var = "AGE",
  sdtm_domain = "DM", sdtm_var = "AGE", confidence = 1
)
tmc <- build_trace_model(adam, sdtm, mapping = map)

deriv <- data.frame(
  from_dataset = c("ADAE", "ADTTE"),
  from_var     = c("AAGE", "TAGE"),
  to_dataset   = c("ADSL", "ADAE"),
  to_var       = c("AGE", "AAGE")
)

trace_chains(tmc, adam_derivations = deriv)
```

`ADTTE.TAGE` is three hops from source, but it is fully grounded: the chain
reaches `DM.AGE`. The `trace_depth` argument caps how far the walk follows a
chain, and circular derivations are detected and reported rather than looping
forever.

## Focusing on the problems

A coverage table lists every variable. On a real submission with thousands of
variables, a team wants only the ones that need attention. `trace_problems()`
returns just those: variables not traced to the required level, and, when a
derivation table is given, chains that are circular, cut off, or never reach
SDTM.

```{r problems}
trace_problems(tm2)
```

Passing the derivation table adds chain problems to the same view, so cycles and
ungrounded derivations show up alongside untraced variables:

```{r problems-chains}
trace_problems(tmc, adam_derivations = deriv)
```

The view is built from the same coverage and chain functions, so it stays
correct as a submission grows; the engine is exercised on a 10,000-variable
model in the package tests.
